---
title: "Lab 9-2: Support Vector Machines"
output: html_notebook
---
```{r}
library(tidyverse)
library(e1071)
library(palmerpenguins)
```

## Lab 9-2: Support Vector Machines

## Penguin Lecture Example
```{r}
peng <-  penguins %>% 
  drop_na %>%
  select(-c(island,year)) %>% 
  filter(species!='Gentoo') %>%
  droplevels() %>%
  mutate(index = rownames(.)) %>% 
  .[-c(106,15,68,177),] %>% 
  filter(abs(flipper_length_mm -195) <=10) %>%
  filter(abs(bill_length_mm-45)<=5)

ggplot(peng, aes(x=flipper_length_mm, y = bill_length_mm, color=species)) + geom_point()

```


```{r}
svmpoly <- svm(species ~ flipper_length_mm+bill_length_mm, 
               data = peng, type='C-classification',
               kernel = "polynomial", degree=4, coef0 = 1, cost = 10000)
plot(svmpoly, peng, bill_length_mm ~ flipper_length_mm)
```
```{r}
svmradial <- svm(species ~ flipper_length_mm+bill_length_mm, 
               data = peng, type='C-classification',
               kernel = "radial", gamma = 1, cost = 10000)
plot(svmradial, peng, bill_length_mm ~ flipper_length_mm)
```
## Support Vector Machine

In order to fit an SVM using a non-linear kernel, we once again use the `svm()` function. However, now we use a different value of the parameter `kernel`. To
fit an SVM with a polynomial kernel we use `kernel = "polynomial"`, and to fit an SVM with a radial kernel   we use `kernel = "radial"`.
In the former case we also use the `degree` argument to specify a degree for the polynomial kernel (this is $d$ in (9.22)), and in the latter case we use `gamma` to specify a value of $\gamma$ for the radial basis kernel (9.24).

We first generate some data with a non-linear class boundary, as follows:

```{r chunk16}
set.seed(1)
x <- matrix(rnorm(200 * 2), ncol = 2)
x[1:100, ] <- x[1:100, ] + 2
x[101:150, ] <- x[101:150, ] - 2
y <- c(rep(1, 150), rep(2, 50))
dat <- data.frame(x = x, y = as.factor(y))
```

Plotting the data makes it clear that the class boundary is indeed non-linear:

```{r chunk17}
plot(x, col = y)
```

The data is randomly split into training and testing groups. We then fit the training data using the `svm()` function with a radial kernel and $\gamma=1$:

```{r chunk18}
train <- sample(200, 100)
svmfit <- svm(y ~ ., data = dat[train, ], kernel = "radial",  
    gamma = 1, cost = 1)
plot(svmfit, dat[train, ])
```

The plot shows that the resulting  SVM has a decidedly non-linear  boundary. The `summary()` function can be used to obtain some  information about the SVM fit:

```{r chunk19}
summary(svmfit)
```

We can see from the figure that there are a fair number of training errors in this SVM fit.
If we increase the value of `cost`, we can reduce the number of training errors. However, this comes at the price of a more irregular decision boundary that seems to be at risk of overfitting the data.

```{r chunk20}
svmfit <- svm(y ~ ., data = dat[train, ], kernel = "radial", 
    gamma = 1, cost = 1e5)
plot(svmfit, dat[train, ])
```

We can perform cross-validation using `tune()` to select the best choice of $\gamma$ and `cost` for an SVM with a radial kernel:

```{r chunk21}
set.seed(1)
tune.out <- tune(svm, y ~ ., data = dat[train, ], 
    kernel = "radial", 
    ranges = list(
      cost = c(0.1, 1, 10, 100, 1000),
      gamma = c(0.5, 1, 2, 3, 4)
    )
  )
summary(tune.out)
```

Therefore, the best choice of parameters involves `cost = 1` and `gamma = 0.5`.
We can view the test set predictions for this model by applying the `predict()` function to the data. Notice that to do this we subset the dataframe `dat` using `-train` as an index set.

```{r chunk22}
table(
    true = dat[-train, "y"], 
    pred = predict(tune.out$best.model, newdata = dat[-train, ])
    )
```

$12\,\%$ of test observations are misclassified by this SVM.

## ROC Curves

The `ROCR` package can be used to produce ROC curves such as those in Figures9.10 and 9.11. We first write a short function to plot an ROC curve given a vector containing a numerical score for each observation, `pred`, and a vector containing the class label for each observation, `truth`.

```{r chunk23}
library(ROCR)
rocplot <- function(pred, truth, ...) {
  predob <- prediction(pred, truth)
  perf <- performance(predob, "tpr", "fpr")
  plot(perf, ...)
}
```


SVMs and support vector classifiers output class labels for each observation. However, it is also possible to obtain *fitted values* for each observation, which are the numerical scores used to obtain the class labels. For instance, in the case of a support vector classifier, the fitted value for an observation $X= (X_1, X_2, \ldots, X_p)^T$ takes the form $\hat{\beta}_0 + \hat{\beta}_1 X_1 + \hat{\beta}_2 X_2 + \cdots + \hat{\beta}_p X_p$. For an SVM with a non-linear kernel, the equation that yields the fitted value is given in (9.23). In essence, the sign of the fitted value determines on which side of the decision boundary the observation lies. Therefore, the relationship between the fitted value and the class prediction for a given observation is simple: if the fitted value exceeds zero then the observation is assigned to one class, and if it is less than zero then it is assigned to the other.
In order to obtain the fitted values for a given SVM model fit, we use `decision.values = TRUE` when fitting `svm()`.
Then the `predict()` function will output the fitted values.

```{r chunk24}
svmfit.opt <- svm(y ~ ., data = dat[train, ], 
    kernel = "radial", gamma = 2, cost = 1, 
    decision.values = T)
fitted <- attributes(
    predict(svmfit.opt, dat[train, ], decision.values = TRUE)
  )$decision.values
```

Now we can produce the ROC plot. Note we use the negative of the fitted values so that negative values correspond to class 1 and positive values to class 2.

SVM appears to be producing accurate predictions. By increasing $\gamma$ we can produce a more flexible fit and generate further improvements in accuracy.

```{r chunk26}
rocplot(-fitted, dat[train, "y"], main = "Training Data")
svmfit.flex <- svm(y ~ ., data = dat[train, ], 
    kernel = "radial", gamma = 50, cost = 1, 
    decision.values = T)
fitted <- attributes(
    predict(svmfit.flex, dat[train, ], decision.values = T)
  )$decision.values
rocplot(-fitted, dat[train, "y"], add = T, col = "red")
```


However, these ROC curves are all on the training data. We are really more interested in the level of prediction accuracy on the test data. When we compute the ROC curves on the test data, the model with $\gamma=2$ appears to provide the most accurate results.

```{r chunk27}
fitted <- attributes(
    predict(svmfit.opt, dat[-train, ], decision.values = T)
  )$decision.values
rocplot(-fitted, dat[-train, "y"], main = "Test Data")
fitted <- attributes(
    predict(svmfit.flex, dat[-train, ], decision.values = T)
  )$decision.values
rocplot(-fitted, dat[-train, "y"], add = T, col = "red")
```

## SVM with Multiple Classes

If the response is a factor containing more than two levels, then the `svm()` function will perform multi-class classification using the one-versus-one approach. We explore that setting here by generating a third class of observations.

```{r chunk28}
set.seed(1)
x <- rbind(x, matrix(rnorm(50 * 2), ncol = 2))
y <- c(y, rep(0, 50))
x[y == 0, 2] <- x[y == 0, 2] + 2
dat <- data.frame(x = x, y = as.factor(y))
par(mfrow = c(1, 1))
plot(x, col = (y + 1))
```

We now fit an SVM to the data:

```{r chunk29}
svmfit <- svm(y ~ ., data = dat, kernel = "radial", 
    cost = 10, gamma = 1)
plot(svmfit, dat)
```

The `e1071` library can also be used to perform support vector regression, if the response vector that is passed in to `svm()` is numerical rather than a factor.
